%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{foxsim_users_guide}
---
title: "FoxSim users guide"
author: "Dave Ramsey"
output: pdf_document
---

FoxSim is an **R** package for simulating an invading fox population. The simulation model is based on a cellular automata over a spatial map (raster) with cells of a given resolution.  The model is designed to be fitted to observations using Approximate Bayesian Computational methods and the package includes a sequential Monte Carlo sampler to undertake the ABC computations.  Currently the model is set up to simulate an invading fox population in Tasmania using a habitat raster with 3 km cell resolution.  Habitat cells are coded as (1) - suitable habitat, (0) - unsuitable habitat or (-1) - water (ocean).

The package includes observations of fox carcasses either killed on roads (3) or shot by hunters (1) and 'FoxSim' currently simulates both road killed and hunter killed foxes to facilitate comparison with observed carcasses via ABC estimation

The following presents a minimal guide to getting the model up and running.

The latest version of `FoxSim` is available from **GitHub** and in order to install it as an **R** package, it needs to be built from source.   Building source packages requires two packages to be installed to give **R** access to the required toolchain including a suitable C++ compiler.

First, download and install **Rtools** from http://cran.r-project.org/bin/windows/Rtools

Then install `devtools` (from CRAN).  Once **Rtools** and `devtools` has been installed run the following commands.

`library(devtools)`  
`install_gitub("dslramsey/foxsim")`

assuming the package builds OK, you should be good to go. Now load the `FoxSim` package and read in the fox habitat and road maps for Tasmania.

```{r}
library(FoxSim, quietly=T)
data(habitatvec)
data(roadvec)

```
Both `habitatvec` and `roadvec` are Vector objects but each contain an attribute `dims` containing the dimensions of the original raster matrix from which they were created.  This can be accessed by

```{r}
nd<- attributes(habitatvec)$dims
nd
```

which shows the original raster had dimensions of 114 rows and 106 columns.  Now read in the carcass observations.
```{r}
data(carcass_obs)
carcass_obs
```

Here the vector `xr` contains the number of road killed foxes discovered each year starting from 2001 and ending in 2014 (a little presumptuous as 2014 is only a little more than half over) and `xs` contains the number of hunter killed carcasses over the same period.


To run the `foxsim` cellular automata model we must also specifiy two dispersal kernels.  These are used to define two possible dispersal modes and are randomly selected from in the model with equal probability.  Here we will specifiy both a **local** and **vagrant** dispersal modes with a maximum dispersal distance of 30 km.  These are then place in a list for input into `foxsim`.

`TODO` modify C++ code to take arbitrary number of dispersal kernels

```{r}
# local neighbourhood dispersal kernel
kern1<- kernel2D(eps=3, kfun="gamma", shape=1, scale=3, max.disp=30)

# Vagrant (searching) dispersal kernel. equal dispersal probability  
# in all directions out to 30 km.
kern2<- kernel2D(eps=3, kfun="circle", sigma=30, max.disp=30)

kern<- list(kern1, kern2)

kdim<- dim(kern1)[1] # save dimensions of kernel

```

We can visualise these kernels using the following 

```{r,fig.width=10,fig.height=8}
par(mfrow=c(1,2),pty="s")
x<- y<- seq(-30, 30, 3)
image(x, y, kern1, col=terrain.colors(50),main="local dispersal kernel")
image(x, y, kern2, col=terrain.colors(50),main="vagrant dispersal kernel")

```

Finally, we require some introduction points for foxes in the model to begin the invasion process.  The rumoured introductions were said to have occurred at Longford, Oatlands and St Helens.   We will introduce foxes simultaneously at these 3 points and allow for some uncertainty in the actual release locations by randomly selecting a habitat cell within 20 km of each of these locations to use as the release cell.  Cells satisfying these conditions are available in the data file `incpoints`

```{r}
data(incpoints)
incpoints
```

which is a list of length 3 containing a vector of cell numbers within 20k of each location.  The `foxsim` model will randomly select one of these cells from each of the three locations to use as the release point for foxes.

Now we are almost ready to run the model.  The final requirements of the `foxsim` model are some parameters governing the population and obervation processes as well as start and end times. These need to be supplied in a `list` with specific name attributes and we will construct some dummy parameters here to illustrate.

```{r}
Parms<- list(syear=1998,eyear=2013,psurv=0.7,Ryear=1.1,proad=0.2,pshot=0.1)

# Now do a single run of the model

foxsim(nd[1], nd[2], kdim, habitatvec, roadvec, incpoints, kern, Parms)
       
```

`foxsim` produces a list containing 3 vectors of length `length(syear:eyear)`.  The first element of the list contains the number of road killed fox carcasses each year predicted by the model.  The second element contains the predicted hunter killed carcasses each year and the third contains the predicted number of occupied cells each year.

`TODO`  change C++ code to give the list elements names attributes.

While this is all well and good, what we would really like to do is obtain estimates of the parameters in `Parms` by fitting the model to the observed fox carcass data.  We will use a sequential Monte Carlo (SMC) algorithm to fit the model by matching simulated fox carcass predictions with the observed carcass data within an ABC framework.

The SMC algorithm is contained in the function `PMC.sampler` (for population Monte Carlo).  Its required arguments are `N` the number of particles to estimate, where an individual particle refers to a vector of parameters e.g. `Parms`, `x0` which is the list of observation vectors, `SeqTol` is a vector containing the decreasing list of tolerance values used for ABC matching, `priors` is a list of specifications for the prior distributions of the parameters and `Data` is a list containing the other fixed data that is required by `foxsim`.

First we construct the fixed data.  This contains all the info required by `foxsim` in a list and can be constructed directly

```{r}
MyData<- list(nr=nd[1],nc=nd[2],kdim=kdim, hab.vec=habitatvec, road.vec=roadvec,  
              ipoints=incpoints, kern.list=kern, eyear=2014)

```

Since the end year of the observations is fixed at 2014, we place it in `MyData` as it is not a parameter to be estimated.  However the fox introduction year `syear` is unknown and is allowed to be estimated by the model.  Here we allow `syear` to take values between 1995 and 2001.

We now need some prior distributions for the parameters in the model.  These are listed in the order introduction year `syear`, cell survival `psurv`, road kill probability `proad`, hunter kill probability `pshot` and cell reproductive rate `Ryear`.


```{r}
priors<- list(list("uniform",-3.49,3.49),
              list("beta",1,1),
              list("beta",17,83),
              list("beta",17,83),
              list("uniform",0.1,2))
```

Here we transform the prior for `syear` to be a number on the real line between -3.49, 3.49 with 0 corresponding to the year 1998.  This facilitates easier sampling by the SMC algorithm with any value proposed by the model, say `val`, transformed into the integer number for `syear` using `round(val + 1998)`.

Finally we can now do some SMC updating by specifying some tolerances and running the following commands

```{r}
Tol<- c(5,4)

fit<- PMC.sampler(N=10, carcass_obs, Tol, priors, MyData, parallel=F)
theta<- fit$Tol4$theta  # parameter estimates for final tolerance
theta[,1]<- round(theta[1] + 1998) # transform syear to integer year of intro.
theta

```


